Typing a linear \picalc

### 2020-03-09 paper/paper.tex
trying to see where one figure ends and the other begins...
Add draft on syntax

variable references (strings, locally named, de Bruijn)

allows to ignore alpha conversion, or proofs of inequality between strings

strings to maybe de Bruijn, names can be kept in context as well, just not doing it

@@

  \inferrule
  {n : \N}
  {0 : \Var_n}

  \inferrule
  {x : \Var_n}
  {\suc x : \Var_{\suc n}}
\end{mathpar}
\caption{Types of size $n$}
\end{figure}

@@

\begin{equation*} \label{pi-calc-syntax}
\begin{split}
  \Process_n ::=& \; \PO_n                    \\
              |& \; \new{} \Process_{\suc n}          \\
              |& \; \comp{\Process_n}{\Process_n}          \\
              |& \; \recv{\Var_n}{}\Process_{\suc n} \\
              |& \; \send{\Var_n}{\Var_n}\Process_n
\end{split}
\end{equation*}
\caption{Well-scoped grammar using de Bruijn indices}
\end{figure}

@@

  \inferrule
  {\gamma : \PreCtx_n \\ t : \Type}
  {\gamma , t : \PreCtx_{\suc n}}
\end{mathpar}
\end{figure}

@@

  \inferrule
  {\Gamma : \Ctx_{is} \\ x : \Carrier_i}
  {\Gamma , x : \Ctx_{is , i}}
\end{mathpar}
\end{figure}
two-layered approach: types on one hand, capabilities on the other
removing from context vs keeping in context but marking it used

@@

  \inferrule
  {\Gamma : \Ctx_{is} \\ x : \Carrier_i \\ x' : \Carrier_j \\ \Delta : \Ctx_{is} \\\\
    loc_x : \contains{\gamma}{\Gamma}{t}{x}{\Delta}}
  {\constr{suc} : \contains{\gamma , t}{\Gamma , x'}{t}{x}{\Delta , x'}}
\end{mathpar}
\end{figure}

@@

  \inferrule
  {left : \types{\gamma}{\Gamma}{P}{\Delta} \\\\
    right : \types{\gamma}{\Delta}{Q}{\Xi}}
  {\constr{comp} : \types{\gamma}{\Gamma}{\comp{P}{Q}}{\Xi}}
\end{mathpar}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Subject reduction}

>>>>

Abstraction is one of the key reasoning tools in a language: it allows for features to be defined over a range of inputs.
Such is the case of the \picalc{} too, where both scope restriction and input introduce abstractions.
The bodies of these constructs must have a way of referring to their argument.
Using names as variable references is a popular option amongst humans.
However, names are cumbersome to mechanize: inserting a new variable into an environment means proving that the name of such variable is different to all other variable names in the environment.
Moreover, to a machine names are of no significance whatsoever.

Machines prefer things they can algorithmically act upon. Like natural numbers!
What does it mean to use a number as a variable reference?
The idea de Bruijn had \cite{} was to use the index $n$ to refer to the variable introduced $n$ binders ago.
The binders themselves introduce no names anymore.
The expression $\lambda g . (\lambda f . f g) g$ in the $\lambda$-calculus would translate as $\lambda (\lambda 0 1) 0$.
That is, terms at different \emph{depths} must use different indices to refer to the same binding.
Humans find this often confusing.
There is however no reason not to keep the original names bestowed by the humans together with the indices.
Machines can then manipulate references mechanically and still use names to present them to humans.

A variable occurring under $n$ abstractions has $n$ things to refer to.
References outside of that range have no associated meaning.
It is useful to rule out these nonsensical terms syntactically.
In Figure \ref{var} we do so by introducing the indexed family of types $\Var_n$: for all naturals $n$, the type $\Var_n$ has $n$ distinct elements.

@@

  \inferrule
  {n : \N}
  {0 : \Var_{\suc n}}

  \inferrule
  {x : \Var_n}
  {\suc x : \Var_{\suc n}}
\end{mathpar}
\caption{Types of size $n$}
\label{var}
\end{figure}

Every time we go under a binder, the number of binders a variable might refer to increments by one.
To propagate this information, we index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
As shown in Figure \ref{process}, we increase the \emph{depth} counter every time we create a new channel or receive some input.

@@

\begin{equation*} \label{pi-calc-syntax}
\begin{split}
  \Process_n ::=& \; \PO_n                    \\
              |& \; \new{} \Process_{\suc n}          \\
              |& \; \comp{\Process_n}{\Process_n}          \\
              |& \; \recv{\Var_n}{}\Process_{\suc n} \\
              |& \; \send{\Var_n}{\Var_n}\Process_n
\end{split}
\end{equation*}
\caption{Well-scoped grammar using de Bruijn indices}
\label{process}
\end{figure}

Unlike with names, using type-level de Bruijn indices makes our syntax well-scoped by construction.
As a consequence, the semantics of our language can be defined on the totality of the syntax.
User-friendliness can still be recovered through a function that converts processes with names into (possibly) processes with indices.
This function would keep track of what index is associated with what name, and would traverse the process recursively, taking note of new binders and substituting variable references.
If the process is ill-scoped, the function would return nothing.
To print things back to the user names can be substituted with an index together with the name.

@@

  \inferrule
  {\gamma : \PreCtx_n \\ t : \Type}
  {\gamma , t : \PreCtx_{\suc n}}
\end{mathpar}
\caption{This is...}
\end{figure}

@@

  \inferrule
  {\Gamma : \Ctx_{is} \\ x : \Carrier_i}
  {\Gamma , x : \Ctx_{is , i}}
\end{mathpar}
\caption{This is...}
\end{figure}
two-layered approach: types on one hand, capabilities on the other
removing from context vs keeping in context but marking it used

@@

  \inferrule
  {\Gamma : \Ctx_{is} \\ x : \Carrier_i \\ x' : \Carrier_j \\ \Delta : \Ctx_{is} \\\\
    loc_x : \contains{\gamma}{\Gamma}{t}{x}{\Delta}}
  {\constr{suc} : \contains{\gamma , t}{\Gamma , x'}{t}{x}{\Delta , x'}}
\end{mathpar}
\caption{This is...}
\end{figure}

@@

  \inferrule
  {left : \types{\gamma}{\Gamma}{P}{\Delta} \\\\
    right : \types{\gamma}{\Delta}{Q}{\Xi}}
  {\constr{comp} : \types{\gamma}{\Gamma}{\comp{P}{Q}}{\Xi}}
\end{mathpar}
\caption{This is...}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Subject reduction}
